Good Empirics Dominate Bad Theory: Comments on Blair and Sambanis, 2020, \"Forecasting Civil Wars: Theory and Structure in an Age of 'Big Data' and Machine Learning\", *JCR*

### 2020-06-04 paper.Rmd
moar mdw
changes
add local fallback .bib file
so that Rick can knit the file; we will need this eventually anyways for reproducability

---
title: "Good Empirics Dominate Bad Theory: Comments on Blair and Sambanis, 2020, \"Forecasting Civil Wars: Theory and Structure in an Age of 'Big Data' and Machine Learning\", *JCR*"
author: Richard K. Morgan and Andreas Beger and Michael D. Ward
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  pdf_document:
    keep_tex: true
references:
- id: blair2020forecasting
  title: "Forecasting Civil Wars: Theory and Structure in an Age of 'Big Data' and Machine Learning"
  author:
  - family: Blair
    given: Robert A.
  - family: Sambanis
    given: Nicholas
  container-title: Journal of Conflict Resolution
  type: article-journal
  issued:
    year: 2020
bibliography: /Users/mdw/Git/whistle/master.bib
---

@@

# Review of Blair and Sambanis 2020

@@

Lastly, the AUC-ROC values reported in the original BS Table 2 are calculated on the basis of slightly different numbers of underlying test set cases (see Table \ref{table2-N}). ROC calculations for a set of predictions can only be done on the set of cases for which both non-missing predictions and non-missing outcomes are available. Those sets differ across models (columns) for each row in Table 2. Thus a difference in AUC-ROC values for two models could be due to the fact that they were calculated on different sets of underlying cases, not because the models are systemically performing at a different level. In other words, the results for different models in BS Table 2 are actually not comparable to the other, and any conclusions drawn from such comparison are potentially incorrect.

@@

Table \ref{table1-smooth} is our replication of BS Table 1 with smoothed AUC-ROC. The results differ slightly from the original BS Table 1, typically by no more than 0.01, due to the non-deterministic nature of the RF models. It is the case that BS set the RNG seed in their replication code, which should theoretically allow exact reproduction, but (1) there was a change in more recent versions of R that affected the RNG seeding process, and (2) we refactored the replication script to allow one to run the models in parallel. In any case, the interpretation of results should not be sensitive to random variation, i.e. it should not depend on using a specific RNG seed. On the basis of these results, BS conclude that the Escalation model is generally superior to the alternatives, and we can replicate that interpretation when using smoothed ROC curves.

@@

Table \ref{table1-nosmooth} shows a version of BS Table 1 with the conventional non-smoothed ROC curves. The Average model outperforms the Escalation model in 17 out of 18 cases, and the CAMEO model outperforms in 16 of 18 cases, with one tie. The Goldstein model generally outperforms the Escalation model in the 6-month version. The Quad model appears to be roughly on par with the Escalation model. Thus the original BS conclusion that the Escalation model is superior to the alternative models is completely conditional on the non-standard use of smoothed ROC curves, and overturns when using traditional AUC-ROC calculations.

@@

Table \ref{table2-fixed} shows our replication of BS Table 2 with (1) regular, not smoothed, AUC-ROC, (2) fixed "Weighted by PITF" and "PITF Split-Population" models, and (3) AUC-ROC values computed on the common joint subset of tests cases that all models have non-missing predictions for. Table \ref{table2-full} further below shows AUC-ROC values for both smoothed and non-smoothed versions, and both the original, model-varying test cases sets and our common joint subset.

@@

What impact did the ROC smoothing have overall on the results reported in BS Tables 1 and 2? Tables \ref{table1-benefit} and \ref{table2-benefit} show the increase in AUC-ROC when using smoothed ROC curves, compared to the standard non-smoothed AUC-ROC. Positive values indicate that smoothing increased a model's apparent performance. The Escalation model is the only model that consistently had a benefit from smoothing. For all 8 other models, smoothing sometimes gave a benefit, sometimes not.

>>>>

---
title: "Good Empirics Dominate Bad Theory: Comments on Blair and Sambanis, 2020, \"Forecasting Civil Wars: Theory and Structure in an Age of 'Big Data' and Machine Learning\", *JCR*"
author: Richard K. Morgan and Andreas Beger and Michael D. Ward
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  pdf_document:
    keep_tex: true
references:
- id: blair2020forecasting
  title: "Forecasting Civil Wars: Theory and Structure in an Age of 'Big Data' and Machine Learning"
  author:
  - family: Blair
    given: Robert A.
  - family: Sambanis
    given: Nicholas
  container-title: Journal of Conflict Resolution
  type: article-journal
  issued:
    year: 2020
bibliography: 
  - "`r user <- Sys.info()[['user']]; if (user=='andybega')  '/Users/andybega/Work/whistle/master.bib' else if (user=='mdw') '/Users/mdw/Git/whistle/master.bib' else references.bib`"
---

@@

@rost:etal:2009,  GLOBUS, 
# Review of Blair and Sambanis 2020

@@

Lastly, the AUC-ROC values reported in the original BS Table 2 are calculated on the basis of slightly different numbers of underlying test set cases (see Table \ref{tab:table2-N}). ROC calculations for a set of predictions can only be done on the set of cases for which both non-missing predictions and non-missing outcomes are available. Those sets differ across models (columns) for each row in Table 2. Thus a difference in AUC-ROC values for two models could be due to the fact that they were calculated on different sets of underlying cases, not because the models are systemically performing at a different level. In other words, the results for different models in BS Table 2 are actually not comparable to the other, and any conclusions drawn from such comparison are potentially incorrect.

@@

Table \ref{tab:table1-smooth} is our replication of BS Table 1 with smoothed AUC-ROC. The results differ slightly from the original BS Table 1, typically by no more than 0.01, due to the non-deterministic nature of the RF models. It is the case that BS set the RNG seed in their replication code, which should theoretically allow exact reproduction, but (1) there was a change in more recent versions of R that affected the RNG seeding process, and (2) we refactored the replication script to allow one to run the models in parallel. In any case, the interpretation of results should not be sensitive to random variation, i.e. it should not depend on using a specific RNG seed. On the basis of these results, BS conclude that the Escalation model is generally superior to the alternatives, and we can replicate that interpretation when using smoothed ROC curves.

@@

Table \ref{tab:table1-nosmooth} shows a version of BS Table 1 with the conventional non-smoothed ROC curves. The Average model outperforms the Escalation model in 17 out of 18 cases, and the CAMEO model outperforms in 16 of 18 cases, with one tie. The Goldstein model generally outperforms the Escalation model in the 6-month version. The Quad model appears to be roughly on par with the Escalation model. Thus the original BS conclusion that the Escalation model is superior to the alternative models is completely conditional on the non-standard use of smoothed ROC curves, and overturns when using traditional AUC-ROC calculations.

@@

Table \ref{tab:table2-fixed} shows our replication of BS Table 2 with (1) regular, not smoothed, AUC-ROC, (2) fixed "Weighted by PITF" and "PITF Split-Population" models, and (3) AUC-ROC values computed on the common joint subset of tests cases that all models have non-missing predictions for. Table \ref{tab:table2-full} further below shows AUC-ROC values for both smoothed and non-smoothed versions, and both the original, model-varying test cases sets and our common joint subset.

@@

What impact did the ROC smoothing have overall on the results reported in BS Tables 1 and 2? Tables \ref{tab:table1-benefit} and \ref{tab:table2-benefit} show the increase in AUC-ROC when using smoothed ROC curves, compared to the standard non-smoothed AUC-ROC. Positive values indicate that smoothing increased a model's apparent performance. The Escalation model is the only model that consistently had a benefit from smoothing. For all 8 other models, smoothing sometimes gave a benefit, sometimes not.
